### 2014-04-29 doc/ApTranscriptome_manuscript.Rmd
Updated gitignore. Working on manuscript

We apply this approach in a study of two closely related eastern North American ant species, *Aphaenogaster rudis* and *A. picea* [@umphrey1996], which have more southern and northern distributions, respectively. Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguezcabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006]. As ectothermic species, ants experience a wide range of temperature conditions. Throughout the year, ant species persist through freezings and hot summers, and at shorter time scales, individual workers can experience substantial thermal shock as they leave the buffered ant nest to forage for food. Thus, at the organismal level we expect that ants utilize multiple molecular pathways to tolerate thermal stress.

@@

While characterizing the reactionome is useful for description of the molecular pathways for thermal tolerance, a greater understanding of the process of evolution can be gained using a comparative approach [@harvey1991]. Specifically, comparisons of populations or species in contrasting environments can reveal the extent to which geographic differentiation and selection has changed the number of genes involved in thermal tolerance. To accomplish this, we compared the thermal reactionomes between the two species of *Aphaenogaster*. *A. picea*, the northern species, experiences less chronic and acute heat stress than *A. rudis*. Thus, we that *A. picea* would have more genes than *A. rudis* that are up-regulated to high temperatures, while *A. rudis* would have higher constitutive expression of these thermally-tolerant genes. Reciprocally, we expected *A. rudis* to have greater up-regulation of genes in response to cold temperatures, while *A. picea* would have greater constitutive expression of cold-tolerant genes. Finally, phenotypic studies in *Aphaenogaster* [@warren2013] and other ecotherms [@hoffmann2013] have found less variation in critical maximum than minimum temperatures. From this, we hypothesized that if the species had more similar maximum temperatures than minimum temperatures, the northern species *A. picea* would have a broader thermal breadth of gene expression than *A. rudis*.

@@

*Aphaenogaster picea* was collected from Molly Bog, Vermont (University of Vermont Natural Areas; lat, lon) and *A. rudis* was collected from Durham, North Carolina ( 36.037° N -78.874° W), designated as *A. picea* and *A. rudis* henceforth. Though the phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], preliminary work supports that these species are separate species (Bernice Bacon DeMarco, pers. communication). species of these species were maintained in the lab for 5 months prior to experimentation.

From a single species of each species, we haphazardly collected 12 ants at the same time on 12 days. Each day, the 12 ants were placed in glass tubes into a water bath (...) at one of 12 randomly-assigned temperatures, every 3.5°C between 0° and 38.5°C, for one hour. The minimum and maximum temperatures were selected based on preliminary work showing that these temperatures are close to the critical minimum and maximum temperatures for an *Aphaenogaster* [@warren2013. At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C. mRNA was extracted from three pooled ants using a DNeasy kit (Qiagen Inc; Valencia, CA). Ants were homogenized in Buffer ATL with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY) and the standard Qiagen protocol was followed from this point.

@@

For each species, the 12 samples were barcoded and sequenced in a single lane of 2 x 100bp paired-end reads on an Illumina HiSeq 1500 yielding 200 and 160 million reads for the A. picea and A. rudis samples respectively. Reads were filtered to remove Illumina adapter sequences and low quality bases using the program Trimmomatic [@lohse2012].

@@

With the same set of genes, we explored the extent to which differences in the thermal reactionome between the species were due to changes in the mean or shape of the reaction norm for each individual gene. Following the example of Murren et al [-@murren2014], we tested for overall differences in the (a) mean expression value across temperatures (*M*), (b) slope of expression (*S*), (c) curvature of expression (*C*) and (d) wiggle of expression (*W*), which consists of all higher-order differences in shape not capture by the first three measures (formulas given in Appendix). As genes differed by five orders of magnitude in expression levels, we calculated the mean-standardized difference between the species $\Delta M = (Mean_{A. picea} - Mean_{Ar}) / \bar M$, where $\bar M = (Mean_{A. picea} + Mean_{Ar}) / 2$ yielding $\Delta M$, $\Delta S$, $\Delta C$, and $\Delta W$ for the mean, slope, curvature and wiggle of each gene, respectively. Using the mean-standardized values, we performed one-sample two-sided *t*-tests to determine if the overall mean, slope or curvature were greater in *A. picea* than *A. rudis*.

@@

To test if genes that were responsive to cold temperatures in the southern *A. rudis* species or hot temperatures in the northern *A. picea* species had higher constitutive expression in the contrasting species, we performed non-parametric two-sided Wilcoxon tests comparing expression values near the optimum of 19.5° C.

As critical maximum temperatures are more constrained than critical minimum temperatures [@warren2013; @hoffman2013], we predicted that the more northern *A. picea* species would have a broader thermal range of stable gene expression than *A. rudis*. To test this with our data, for each *Intermediate* expressed gene in each species, we randomly sampled 1,000 temperature values weighted by the expression function (e.g. point of maximum expression had highest probability of being sampled). We then calculated the standard deviation of this random draw, and performed a *t*-test comparing the standard deviations of expression between species. Conversely, we predicted that the thermal sensitivity of *Bimodal* expressed genes would be greater in *A. rudis* than *A. picea*. We performed the same analysis as with the *Intermediate* transcripts, but with low standard deviation being indicative of a relatively un-sensitive gene (nearly flat), while the maximum standard deviation of expression is for a highly-convex expression pattern. Thus, finding one species had overall greater standard deviation of expression for *Bimodal* genes would be indicative of greater thermal sensitivity.

@@

We identified 22,495 genes with overall model fits of P < 0.05, retaining 9,809 after correcting for FDR. Of these transcripts, 8,044 significantly change expression with temperature including 6,727 that have a species x temperature interaction. The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not expressed differed significantly between the species ($\chi^2 = 170, P < 0.0001$; Table 1). For both species, the majority of responsive transcripts were in the *Low* category (Fig. 1). More transcripts had *Intermediate* expression in *A. rudis* than in *A. picea*, which had nearly twice as many bimodally-expressed transcripts  (Fig. 1, Table 1).

```{r expression_type_table, echo=FALSE, results='asis'}
pandoc.table(Ap.type.table, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each species and their overlap.")
```

@@

![Probability density function of peak expression for all thermally-responsive in *A. rudis* (blue) and *A. picea* (red) species](../results/PDF_expression_all.png)

@@

We quantified differences in the reaction norms of thermally-responsive genes between species as changes to due shifts in mean ($\Delta M$), slope ($\Delta S$), curvature ($\Delta C$) and wiggle ($\Delta W$). Overall, we found that there was no significant difference in position of the mean (*t* = -0.95, *P* = 0.34), but changes in the shape due to slope (*t* = 24.67, *P* < 0.001), curvature (*t* = 15.71, *P* < 0.001) and wiggle (*t* = -12.45, *P* < 0.001) were all highly significant. On average, slope (95% CI 0.41 - 0.48) and curvature (95% CI 0.25 - 0.33) were greater in *A. picea* than *A. rudis*, indicating a steeper response to temperature.

@@

Plasticity is a well-characterized mechanism by which organism tolerate environmental change. Given our sampling of a warmer (*A. rudis*) and colder (*A. picea*) climate species, we hypothesized that genes which are plastically-expressed under heat shock in the colder species would have greater constitutive expression in the warmer species, and vice versa. Consistent with one of our predictions, we found that constitutive expression was greater in the more southern *A. rudis* species for *High* genes in *A. picea* (Wilcoxon V = `r w1$statistic`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r w1$conf.int[1:2]`). However, for genes with *Low* expression in *A. rudis* we also found that constitutive expression was greater in *A. rudis* than *A. picea* (Wilcoxon V = `r w2$statistic`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r w2$conf.int[1:2]`).

We tested for a transcriptome-wide pattern of thermal tolerance and sensitivity by examining the thermal range) for the *Intermediate* expressed genes. Contrary to our expectations based on critical maximum and minimal temperatures, *A. rudis* had a transriptome-wide pattern of broader thermal range (*t*~df=1406~ = 9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64). In contrast, there was no difference in thermal sensitivity as measured by the standard deviation of expression for *Bimodal* expressed genes (*t*~df=1406~ = 0.31, *P* = 0.76).

We performed gene set enrichment analysis on the set difference of each expression category to reveal differences among the species response to thermal shock. For *High* genes, there were 4 enriched GO terms in *A. picea* and 12 in *A. rudis*. For *Low* genes, there were 40 enriched GO terms in *A. picea* and 23 in *A. rudis*. For *Intermediate* genes, there were 3 enriched GO terms in *A. picea* and 54 in *A. rudis*. For *Bimodal* genes, there were 12 enriched GO terms in *A. picea* and 10 in *A. rudis*.

@@

Thermal tolerance is a key trait for determining the survival of individual organisms, the persistence of populations and the geographic limits of species. While considerable research has focused on a few key genes (e.g. heat shock proteins) involved in thermal stress response [...], little is known about the functional diversity or transcriptome-wide patterns of thermal responsiveness. In this study, we characterized the thermal reactionome of sister ant species, *Aphaenogaster picea* and *Aphaenogaster rudis*. We found that nearly 10% of the genes in this species show significant thermal responsiveness.

>>>>

We apply this approach in a study of two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996], which have more southern and northern distributions, respectively. Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguezcabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006]. As ectothermic species, ants experience a wide range of temperature conditions. Throughout the year, ant species persist through freezings and hot summers, and at shorter time scales, individual workers can experience substantial thermal shock as they leave the buffered ant nest to forage for food. Thus, at the organismal level we expect that ants utilize multiple molecular pathways to tolerate thermal stress.

@@

While characterizing the reactionome is useful for description of the molecular pathways for thermal tolerance, a greater understanding of the process of evolution can be gained using a comparative approach [@harvey1991]. Specifically, comparisons of populations or species in contrasting environments can reveal the extent to which geographic differentiation and selection has changed the number of genes involved in thermal tolerance. To accomplish this, we compared the thermal reactionomes between the two species of *Aphaenogaster*. *A. picea*, the northern species, experiences less chronic and acute heat stress than *A. carolensis*. Thus, we predicted that *A. picea* would have more genes than *A. carolinensis* that are up-regulated to high temperatures, while *A. carolinensis* would have higher constitutive expression of these thermally-tolerant genes. Reciprocally, we expected *A. carolinensis* to have greater up-regulation of genes in response to cold temperatures, while *A. picea* would have greater constitutive expression of cold-tolerant genes. Finally, phenotypic studies in *Aphaenogaster* [@warren2013] and other ecotherms [@hoffmann2013] have found less variation in critical maximum than minimum temperatures. From this, we hypothesized that if the species had more similar maximum temperatures than minimum temperatures, the northern species *A. picea* would have a broader thermal breadth of gene expression than *A. carolinensis*.

@@

*Aphaenogaster picea* was collected from Molly Bog, Vermont (University of Vermont Natural Areas; lat, lon) and *A. carolinensis* was collected from Durham, North Carolina ( 36.037° N -78.874° W), designated as *A. picea* and *A. carolinensis* henceforth. Though the phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], preliminary work supports that these species are separate species (Bernice Bacon DeMarco, pers. communication). species of these species were maintained in the lab for 5 months prior to experimentation.

From a single species of each species, we haphazardly collected 12 ants at the same time on 12 days. Each day, the 12 ants were placed in glass tubes into a water bath (...) at one of 12 randomly-assigned temperatures, every 3.5°C between 0° and 38.5°C, for one hour. The minimum and maximum temperatures were selected based on preliminary work showing that these temperatures are close to the critical minimum and maximum temperatures for an *Aphaenogaster* [@warren2013. At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C. mRNA was extracted from three pooled ants using a RNeasy kit (Qiagen Inc; Valencia, CA). Ants were homogenized in Buffer ATL with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY) and the standard Qiagen protocol was followed from this point.

@@

For each species, the 12 samples were barcoded and sequenced in a single lane of 2 x 100bp paired-end reads on an Illumina HiSeq 1500 yielding 200 and 160 million reads for the A. picea and A. carolinensis samples respectively. Reads were filtered to remove Illumina adapter sequences and low quality bases using the program Trimmomatic [@lohse2012].

@@

With the same set of genes, we explored the extent to which differences in the thermal reactionome between the species were due to changes in the mean or shape of the reaction norm for each individual gene. Following the example of Murren et al [-@murren2014], we tested for overall differences in the (a) mean expression value across temperatures (*M*), (b) slope of expression (*S*), (c) curvature of expression (*C*) and (d) wiggle of expression (*W*), which consists of all higher-order differences in shape not capture by the first three measures (formulas given in Appendix). As genes differed by five orders of magnitude in expression levels, we calculated the mean-standardized difference between the species $\Delta M = (Mean_{A. picea} - Mean_{Ar}) / \bar M$, where $\bar M = (Mean_{A. picea} + Mean_{Ar}) / 2$ yielding $\Delta M$, $\Delta S$, $\Delta C$, and $\Delta W$ for the mean, slope, curvature and wiggle of each gene, respectively. Using the mean-standardized values, we performed one-sample two-sided *t*-tests to determine if the overall mean, slope or curvature were greater in *A. picea* than *A. carolinensis*.

@@

To test if genes that were responsive to cold temperatures in the southern *A. carolinensis* species or hot temperatures in the northern *A. picea* species had higher constitutive expression in the contrasting species, we performed non-parametric two-sided Wilcoxon tests comparing expression values near the optimum of 19.5° C.

As critical maximum temperatures are more constrained than critical minimum temperatures [@warren2013; @hoffman2013], we predicted that the more northern *A. picea* species would have a broader thermal range of stable gene expression than *A. carolinensis*. To test this with our data, for each *Intermediate* expressed gene in each species, we randomly sampled 1,000 temperature values weighted by the expression function (e.g. point of maximum expression had highest probability of being sampled). We then calculated the standard deviation of this random draw, and performed a *t*-test comparing the standard deviations of expression between species. Conversely, we predicted that the thermal sensitivity of *Bimodal* expressed genes would be greater in *A. carolinensis* than *A. picea*. We performed the same analysis as with the *Intermediate* transcripts, but with low standard deviation being indicative of a relatively un-sensitive gene (nearly flat), while the maximum standard deviation of expression is for a highly-convex expression pattern. Thus, finding one species had overall greater standard deviation of expression for *Bimodal* genes would be indicative of greater thermal sensitivity.

@@

We identified 22,495 genes with overall model fits of P < 0.05, retaining 9,809 after correcting for FDR. Of these transcripts, 8,044 significantly change expression with temperature including 6,727 that have a species x temperature interaction. The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not expressed differed significantly between the species ($\chi^2 = 170, P < 0.0001$; Table 1). For both species, the majority of responsive transcripts were in the *Low* category (Fig. 1). More transcripts had *Intermediate* expression in *A. carolinensis* than in *A. picea*, which had nearly twice as many bimodally-expressed transcripts  (Fig. 1, Table 1).

```{r expression_type_table, echo=FALSE, results='asis'}
pandoc.table(tt2, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each species and their overlap.")

mosaicplot(tt2, shade = FALSE, xlab = "ApVT", ylab = "AcNC", main = "", cex = 0.5)
```

@@

![Probability density function of peak expression for all thermally-responsive in *A. carolinensis* (blue) and *A. picea* (red) species](../results/PDF_expression_all.png)

@@

We quantified differences in the reaction norms of thermally-responsive genes between species as changes to due shifts in mean ($\Delta M$), slope ($\Delta S$), curvature ($\Delta C$) and wiggle ($\Delta W$). Overall, we found that there was no significant difference in position of the mean (*t* = -0.95, *P* = 0.34), but changes in the shape due to slope (*t* = 24.67, *P* < 0.001), curvature (*t* = 15.71, *P* < 0.001) and wiggle (*t* = -12.45, *P* < 0.001) were all highly significant. On average, slope (95% CI 0.41 - 0.48) and curvature (95% CI 0.25 - 0.33) were greater in *A. picea* than *A. carolinensis*, indicating a steeper response to temperature.

@@

Plasticity is a well-characterized mechanism by which organism tolerate environmental change. Given our sampling of a warmer (*A. carolinensis*) and colder (*A. picea*) climate species, we hypothesized that genes which are plastically-expressed under heat shock in the colder species would have greater constitutive expression in the warmer species, and vice versa. Consistent with one of our predictions, we found that constitutive expression was greater in the more southern *A. carolinensis* species for *High* genes in *A. picea* (Wilcoxon V = `r w1$statistic`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r w1$conf.int[1:2]`). However, for genes with *Low* expression in *A. carolinensis* we also found that constitutive expression was greater in *A. carolinensis* than *A. picea* (Wilcoxon V = `r w2$statistic`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r w2$conf.int[1:2]`).

We tested for a transcriptome-wide pattern of thermal tolerance and sensitivity by examining the thermal range) for the *Intermediate* expressed genes. Contrary to our expectations based on critical maximum and minimal temperatures, *A. carolinensis* had a transriptome-wide pattern of broader thermal range (*t*~df=1406~ = 9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64). In contrast, there was no difference in thermal sensitivity as measured by the standard deviation of expression for *Bimodal* expressed genes (*t*~df=1406~ = 0.31, *P* = 0.76).

We performed gene set enrichment analysis on the set difference of each expression category to reveal differences among the species response to thermal shock. For *High* genes, there were 4 enriched GO terms in *A. picea* and 12 in *A. carolinensis*. For *Low* genes, there were 40 enriched GO terms in *A. picea* and 23 in *A. carolinensis*. For *Intermediate* genes, there were 3 enriched GO terms in *A. picea* and 54 in *A. carolinensis*. For *Bimodal* genes, there were 12 enriched GO terms in *A. picea* and 10 in *A. carolinensis*.

@@

Thermal tolerance is a key trait for determining the survival of individual organisms, the persistence of populations and the geographic limits of species. While considerable research has focused on a few key genes (e.g. heat shock proteins) involved in thermal stress response [...], little is known about the functional diversity or transcriptome-wide patterns of thermal responsiveness. In this study, we characterized the thermal reactionome of sister ant species, *Aphaenogaster picea* and *Aphaenogaster carolinensis*. We found that nearly 10% of the genes in this species show significant thermal responsiveness.
